Automatic differentiation via off-policy particle filters

### 2024-03-06 mop/mop.tex
Updates from Overleaf
merge

Bias in dropping the resampling gradient increases as process mixes slower. ADPF has a long memory, and the variance in the weights add up, and there is significant instability. However, by the time the system mixes, ancestral trajectories don't matter as much, so we shouldn't bother with estimating gradients that rely on more than one timestep of the conditional log-likelihood.

@@

@@

Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ is properly weighted for $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
To see  (\ref{L2}),
we write the numerator of (\ref{L1}) as
\[
L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
\]
Using Lemma~1, we see this is properly estimated by
\[
L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j},
\]
from which we obtain (\ref{L2}).

>>>>

\kevin{w.r.t. alpha=0:}
Bias in dropping the resampling gradient increases as process mixes slower. ADPF has a long memory, and the variance in the weights add up, and there is significant instability. However, by the time the system mixes, ancestral trajectories don't matter as much, so we shouldn't bother with estimating gradients that rely on more than one timestep of the conditional log-likelihood.

@@

\noindent {\bf Definition 1, Liu}. {\it A random variable $X$ drawn from a distribution $g$ is said to be properly weighted by a weighting function $w(X)$ with respect to the distribution $\pi$ if for any integrable function $h$,
$$
E_g\{h(X) w(X)\}=E_\pi\{h(X)\}
$$

A set of random draws and weights $\left(x^{(j)}, w^{(j)}\right), j=1,2, \ldots$, is said properly weighted with respect to $\pi$ if
$$
\lim _{m \rightarrow \infty} \frac{\sum_{j=1}^m h\left(x^{(j)}\right) w^{(j)}}{\sum_{j=1}^m w^{(j)}}=E_\pi(h(X))
$$
for any integrable function $h$. In a practical sense we can think of $\pi$ as being approximated by the discrete distribution supported on the $x^{(j)}$ with probabilities proportional to the weights $w^{(j)}$.
}

@@

Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ is properly weighted for $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
To see  (\ref{L2}),
we write the numerator of (\ref{L1}) as
\[
L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
= L^\phi_n \sum_{j=1}^J w_{n,j}^{FC,\theta} \frac{g^\phi_{n,j}}{L_n^\phi}
\]
Using Lemma~1, we resample according to probabilities $\frac{g^\phi_{n,j}}{L_n^\phi}$ to see this is properly estimated by
\[
L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j},
\]
from which we obtain (\ref{L2}).
